---
title: "Lab 1: Eurovision Data Analysis"
author: "Mahmoud Salaymeh and Cesare Raad"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(data.table)
library(rworldmap) 
library(ggthemes)
library(reshape2)
library(e1071) 
library(readxl)
library(scales)
library(knitr)

# Load datasets
url_base <- "https://github.com/Spijkervet/eurovision-dataset/releases/download/2023/"
contestants <- read.csv(paste0(url_base, "contestants.csv"))
votes <- read.csv(paste0(url_base, "votes.csv"))
# Ensure this file is in your working directory if needed
# GDP_per_capita <- read.csv("GDP.PerCapita.WorldCountries.csv")
```

## Question 1: Data Exploration

### A
In this section, we conduct an initial exploration of the Eurovision datasets. The `head()` function allows us to view the first few records to understand the data's format, while the `str()` (structure) function provides a technical summary of the variables, including their data types (numeric, character, etc.) and the overall dimensions of the dataframe. This step is crucial for identifying potential data cleaning needs before proceeding with the analysis.

#### Contestants Data
```{r A.1_contestants,echo=FALSE}
head(contestants,5) %>% select(year,points_final,points_tele_final,points_jury_final,place_final,to_country,song) %>% 
 kable(caption = "First 5 rows of Contestants dataset")
```

#### Votes Data
```{r A.1_votes}
head(votes, 5) %>% 
kable(caption = "First 5 rows of Votes dataset")
str(votes)
```

---

### B.

In this step, we filter the dataset to display Israel's 10 most recent appearances in the Eurovision final, excluding years where they did not qualify. We sort the data by year in descending order.

```{r 1.b_israel_recent}
israel_recent <- contestants %>% 
  filter(to_country_id == "il", !is.na(place_final)) %>%
  arrange(desc(year)) %>% 
  slice_head(n = 10)  %>%
  select(year, song, performer, place_final, points_final)

israel_recent %>% 
  kable(caption = "Israel's 10 Most Recent Eurovision Final Appearances")
```

Based on this 10-appearance window, Israel's best result was achieving 1st place in 2018 with the song "Toy" performed by Netta Barzilai.

### c.

First, we reshape the points data from 2016 onwards into a long format using `pivot_longer` to separate the point sources (total, jury, and tele-vote). 

```{r 1.c_pivot}
contestants_long <- contestants %>%
  filter(year >= 2016) %>%
  select(year, to_country, song, points_final, points_jury_final, points_tele_final) %>%
  pivot_longer(cols = starts_with("points_"), 
               names_to = "Source", 
               values_to = "Points")


head(contestants_long, 6) %>%
  kable(caption = "First 6 rows of the long-format table (2016+)")
```

Next, for the year 2023, we compute the difference between jury points and tele-vote points to see which entries were the most polarizing between the two groups of voters.

```{r 1.c_diff}
diff_2023 <- contestants %>% 
  filter(year==2023) %>%
  mutate(diff=points_jury_final-points_tele_final) %>%
  select(year,song,to_country,points_jury_final,points_tele_final,diff)%>%
  arrange(desc(abs(diff)))

diff_2023 %>% head(5) %>%
  kable(caption = "Top 5 Entries with Largest Jury/Tele-vote Differences in 2023")
```

Among the 5 entries with the largest absolute differences in 2023, the public tele-vote tended to over-rate crowd favorites significantly more than the juries did, as seen by the massive negative differences for Finland, Norway, and Ukraine. In contrast, the juries tended to give high scores to entries like Estonia and Australia that the public largely ignored.

## Question 2: Country-level performance over time

### a.

In this step, we calculate the total number of points each country has received in the finals across all years. We then visualize the top 15 countries using a horizontal bar chart.

```{r 2.a_total_points, fig.height=6, fig.width=8}
top_15_countries <- contestants %>%
  group_by(to_country) %>%
  summarize(total_points = sum(points_final, na.rm = TRUE)) %>%
  arrange(desc(total_points)) %>%
  slice_head(n = 15)

ggplot(top_15_countries, aes(x = total_points, y = reorder(to_country, total_points))) +
  geom_col(fill = "steelblue") +
  labs(title = "Top 15 Countries by Total Eurovision Points in Finals",
       x = "Total Points",
       y = "Country") +
  theme_classic()
```

Sweden, Italy, and the United Kingdom lead the historical rankings in total points, which reflects not only their strong track records of successful performances but also their consistent participation over many decades. Conversely, countries that joined the contest more recently (or have missed many years) have naturally had fewer opportunities to accumulate such massive cumulative point totals.