---
title: "Lab 1: Eurovision Data Analysis"
author: "Mahmoud Salaymeh and Cesare Raad"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    theme: lumen
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
---

```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(data.table)
library(rworldmap) 
library(ggthemes)
library(reshape2)
library(e1071) 
library(readxl)
library(scales)
library(knitr)
```

```{r, eval=TRUE}
# Load datasets
url_base <- "https://github.com/Spijkervet/eurovision-dataset/releases/download/2023/"
contestants <- read.csv(paste0(url_base, "contestants.csv"))
votes <- read.csv(paste0(url_base, "votes.csv"))
```

## Question 1: Data Exploration

### A.
In this section, we conduct an initial exploration of the Eurovision datasets. The `head()` function allows us to view the first few records to understand the data's format, while the `str()` (structure) function provides a technical summary of the variables, including their data types (numeric, character, etc.) and the overall dimensions of the dataframe. This step is crucial for identifying potential data cleaning needs before proceeding with the analysis.

#### Contestants Data
```{r A.1_contestants, eval=TRUE}
contestants%>%select(year,points_final,points_tele_final,points_jury_final,place_final,to_country,song) %>% str()
head(contestants, 5) 


```
#### Votes Data
```{r A.1_votes}
str(votes)
votes %>% head(5)

```
**Data Types Confirmation:** Based on the `str()` output, the key columns (`year`, `points_final`, `points_jury_final`,
         `points_tele_final`, and `place_final`) are indeed numeric (integer or double), and textual columns (`to_country`,
         `song`) are stored as characters.
---

### B.

In this step, we filter the dataset to display Israel's 10 most recent appearances in the Eurovision final, excluding years where they did not qualify. We sort the data by year in descending order.

```{r 1.b_israel_recent}
israel_recent <- contestants %>% 
  filter(to_country_id == "il", !is.na(place_final)) %>%
  arrange(desc(year)) %>% 
  slice_head(n = 10)  %>%
  select(year, song, performer, place_final, points_final)

israel_recent %>% 
  kable(caption = "Israel's 10 Most Recent Eurovision Final Appearances")
```

Based on this 10-appearance window, Israel's best result was achieving 1st place in 2018 with the song "Toy" performed by Netta Barzilai.

### C.

First, we reshape the points data from 2016 onwards into a long format using `pivot_longer` to separate the point sources (total, jury, and tele-vote). 

```{r 1.c_pivot}
contestants_long <- contestants %>%
  filter(year >= 2016) %>%
  select(year, to_country, song, points_final, points_jury_final, points_tele_final) %>%
  pivot_longer(cols = starts_with("points_"), 
               names_to = "Source", 
               values_to = "Points")


head(contestants_long, 6) %>%
  kable(caption = "First 6 rows of the long-format table (2016+)")
```

Next, for the year 2023, we compute the difference between jury points and tele-vote points to see which entries were the most polarizing between the two groups of voters.

```{r 1.c_diff}
diff_2023 <- contestants %>% 
  filter(year==2023) %>%
  mutate(diff=points_jury_final-points_tele_final) %>%
  select(year,song,to_country,points_jury_final,points_tele_final,diff)%>%
  arrange(desc(abs(diff)))

diff_2023 %>% head(5) %>%
  kable(caption = "Top 5 Entries with Largest Jury/Tele-vote Differences in 2023")
```

Among the 5 entries with the largest absolute differences in 2023, the public tele-vote tended to over-rate crowd favorites significantly more than the juries did, as seen by the massive negative differences for Finland, Norway, and Ukraine. In contrast, the juries tended to give high scores to entries like Estonia and Australia that the public largely ignored.

## Question 2: Country-level performance over time

### A.

In this step, we calculate the total number of points each country has received in the finals across all years. We then visualize the top 15 countries using a horizontal bar chart.

```{r 2.a_total_points, fig.height=6, fig.width=8}
top_15_countries <- contestants %>%
  group_by(to_country) %>%
  summarize(total_points = sum(points_final, na.rm = TRUE)) %>% arrange(desc(total_points)) %>%slice_head(n = 15)

ggplot(top_15_countries, aes(x = total_points, y = reorder(to_country, total_points))) +
  geom_col(fill = "steelblue") +
  labs(title = "Top 15 Countries by Total Eurovision Points in Finals",
       x = "Total Points",
       y = "Country") +
  theme_classic()
```

Sweden, Italy, and the United Kingdom lead the historical rankings in total points, which reflects not only their strong track records of successful performances but also their consistent participation over many decades. Conversely, countries that joined the contest more recently (or have missed many years) have naturally had fewer opportunities to accumulate such massive cumulative point totals.

### B.


```{r 2.b, fig.height=6, fig.width=8}
average_points <- contestants %>% group_by(to_country) %>% filter(!is.na(place_final)) %>% summarize(avg_points=mean(points_final,na.rm=TRUE),
          appearances= n()) %>% filter(appearances>=10) %>% arrange(desc(avg_points))

ggplot(average_points,aes(x=avg_points)) + geom_histogram(aes(y=after_stat(density)),
                 bins = 15,binwidth = 15,
                 fill="steelblue",
                 color="black",
                 alpha=0.8) +geom_density(
  linewidth = 1,
  color="black"
) + theme_classic()

Skew_val1<-e1071::skewness(average_points$avg_points)
Kurt_val1<-e1071::kurtosis(average_points$avg_points)


```
```{r 2, fig.height=6, fig.width=8,results='asis'}

cat("Skewness ($\\tilde{\\mu}_3$) = ", round(Skew_val1, 3), "  \n")
cat("Kurtosis ($\\kappa$) = ", round(Kurt_val1, 3), "  \n")

```

**Shape Description:** The distribution is strongly right-skewed (positive skewness), meaning the bulk of countries have relatively low average scores, with a long tail stretching to the right representing the few highly successful outliers. A positive kurtosis indicates the distribution is heavy-tailed, producing extreme values slightly more often than a normal distribution.

### C.

```{r 2.c, fig.height=8, fig.width=10}
Winning_scores <- contestants %>% 
  filter(year >= 1956, !is.na(points_final)) %>% 
  group_by(year) %>% 
  summarize(winning_score = max(points_final, na.rm = TRUE)) 

Winning_scores <- Winning_scores %>% 
  mutate(era = case_when(
    year < 1975 ~ "Pre-1975",
    year >= 1975 & year < 2016 ~ "1975-2015",
    year >= 2016 ~ "2016 Onwards")) %>% 
  group_by(era) %>% 
  mutate(
    era_max = max(winning_score, na.rm = TRUE),
    Normalized = winning_score / era_max
  ) %>% 
  ungroup() %>% 
  select(year, `Score` = winning_score, `Normalized Score` = Normalized) %>% 
  pivot_longer(
    cols = c('Score', 'Normalized Score'),
    names_to = 'Panel',
    values_to = 'score'
  )

ggplot(Winning_scores, aes(x = year, y = score)) + 
  geom_line(color = "gray", linewidth = 1) +
  geom_point(aes(color = score), size = 2.5) +
  scale_color_viridis_c(guide = "none", option = "C") +
  geom_vline(xintercept = 1975, linetype = "dashed", linewidth = 0.8, color = "darkred") + 
  geom_vline(xintercept = 2016, linetype = "dashed", linewidth = 0.8, color = "darkred") +   
  theme_minimal(base_size = 14) + 
  theme(
    strip.background = element_rect(fill = "#e0e0e0", color = NA),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray30")
  ) +
  facet_wrap(~Panel, ncol = 1, scales = "free_y") +
  labs(
    title = "Eurovision Winning Scores Over Time",
    subtitle = "Vertical lines mark voting rule changes in 1975 (12-point system) and 2016 (Jury/Tele-vote split)",
    x = "Year",
    y = "Score"
  )
```

**Normalization:** The scores were normalized by dividing each year's winning score by the maximum winning score within its respective era (Pre-1975, 1975-2015, and 2016 onwards).

**Observations on Rule Changes:** The introduction of the "12 points" system in 1975 caused a visible jump in maximum possible points. In 2016, splitting the jury and tele-vote points essentially doubled the total points available, causing another massive spike in raw scores. However, the normalized panel reveals that the relative dominance of winning songs remains comparable and consistent across all eras despite the underlying point inflation.


## 3
 
### A.
```{r, eval=TRUE}

votes_modern <- votes %>%
  filter(round == "final", year >= 2016, total_points > 0)

top_pairs <- votes_modern %>%
  group_by(from_country_id, to_country_id) %>%
  summarise(
    n_years = n(),
    avg_points = mean(total_points),
    .groups = "drop"
  ) %>%
  filter(n_years >= 5) %>%
  arrange(desc(avg_points)) %>%
  slice_head(n = 15)

print("Top 15 Voting Pairs (2016 onwards):")
print(top_pairs)
```

### B.
```{r}

top_15_countries <- contestants %>%
  filter(year >= 2016) %>%
  group_by(to_country_id) %>%
  summarise(total_received = sum(points_final, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_received)) %>%
  slice_head(n = 15) %>%
  pull(to_country_id)

grid_15 <- expand.grid(
  from_country_id = top_15_countries,
  to_country_id = top_15_countries,
  stringsAsFactors = FALSE
)


avg_given_15 <- votes_modern %>%
  filter(from_country_id %in% top_15_countries, to_country_id %in% top_15_countries) %>%
  group_by(from_country_id, to_country_id) %>%
  summarise(avg_points = mean(total_points), .groups = "drop")


heatmap_data <- grid_15 %>%
  left_join(avg_given_15, by = c("from_country_id", "to_country_id")) %>%
  mutate(
    avg_points = ifelse(from_country_id == to_country_id, NA, avg_points)
  )

overall_mean <- mean(heatmap_data$avg_points, na.rm = TRUE)

ggplot(heatmap_data, aes(x = to_country_id, y = from_country_id, fill = avg_points)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = overall_mean,
    na.value = "grey90", # Colors NA values (including the diagonal) clearly
    name = "Avg Points"
  ) +
  labs(
    title = "Average Points Exchanged Between Top 15 Eurovision Countries (2016+)",
    x = "Receiver (to_country_id)",
    y = "Giver (from_country_id)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```


## 4

### A.
```{r, eval=TRUE}
GDP_per_capita <- read.delim("GDP.PerCapita.WorldCountries.csv",sep="\t")
GDP_per_capita <- GDP_per_capita %>% rename(to_country = Country.Name, gdp_2019=X2019)

contestants_2016_upwards<-contestants %>% filter(year>=2016,!is.na(points_final)) %>% group_by(to_country) %>% summarize(avg_points=mean(points_final),apperances=n()) %>% filter(apperances>= 3)

combined_contestants_gdp<-left_join(contestants_2016_upwards,GDP_per_capita,by="to_country")


ggplot(combined_contestants_gdp,aes(x = gdp_2019,y=avg_points)) +
  geom_point(color="black",size=2)+ geom_text(aes(label = to_country),vjust=-0.8,size = 3)+geom_smooth(method = "lm",se=FALSE,color="green",linetype="solid")+labs(x="GDP Per Capita",y="Average Points",title = "Average Points in Eurovision vs GDP Per Capita for contestants")+theme_classic()

model<- lm(avg_points~gdp_2019,data = combined_contestants_gdp)
summary(model)
```
**Regression Interpretation:** The intercept ($\approx 102.8$) represents the expected average points for a country
with zero GDP. The positive slope ($0.0016$) implies a tiny increase in average points for every unit increase in GDP.However, the $R^2$ is extremely low ($0.096$), indicating GDP explains very little of the variation. Furthermore, the $p$-value ($0.052$) is just above the $0.05$ threshold, meaning GDP per capita is not quite a statistically significant predictor of Eurovision success.
### B.
```{r, eval=TRUE,results='hide'}
country_avg<- contestants %>% filter(!is.na(points_final))%>%group_by(to_country_id)%>%summarize(avg_points=mean(points_final),apperances=n()) %>% filter(apperances>=3) %>% mutate(iso2=toupper(to_country_id))
sPDF <- joinCountryData2Map(country_avg, joinCode = "ISO2",
                            nameJoinColumn = "iso2")
mapCountryData(sPDF, nameColumnToPlot = "avg_points",
               mapRegion = "Eurasia",
               xlim = c(-20, 50), ylim = c(20, 62),
               colourPalette = "heat", catMethod = "pretty",
               addLegend = TRUE,
               mapTitle = "Average Eurovision points per finals appearance")

```

**Geographical Patterns:** The map reveals that Eastern European nations, Scandinavia, and Italy generally score the highest average points, denoted by the hotter colors. Israel is situated at the southeastern extreme of the European map, displaying a moderate historical average compared to the northern powerhouses. Note that Australia is intentionally cut off from this map view due to the applied bounding box.

### C.
 
```{r 4.c_israel_trends, fig.height=8, fig.width=10}
# 1. Filter for Israel's final appearances
 israel_history <- contestants %>% filter(to_country_id=="il",!is.na(place_final)) %>%
  select(year, place_final, points_final) %>%
  pivot_longer(
  cols = c('place_final', 'points_final'),
  names_to = 'metric',
  values_to = 'value') %>%
  mutate(metric = ifelse(metric == "place_final", "Final Placement", "Total Points"))
 
  ggplot(israel_history, aes(x = year, y = value)) +
    geom_line(color = "steelblue", linewidth = 1) +
    geom_point(color = "darkblue", size = 2) +
    geom_vline(xintercept = c(1978, 1979, 1998, 2018), 
               color = "darkgoldenrod", linetype = "dashed", linewidth = 1) +
    theme_minimal(base_size = 14) +
    theme(
      strip.background = element_rect(fill = "#e0e0e0", color = NA),
      strip.text = element_text(face = "bold", size = 12),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, color = "gray30")
    ) +
    facet_wrap(~metric, ncol = 1, scales = "free_y") +
    labs(
      title = "Israel's Eurovision Performance Over Time",
      subtitle = "Dashed gold lines indicate winning years (1978, 1979, 1998, 2018)",
      x = "Year",
      y = ""
    )
```

 
